---
title: Forecasting mass shootings
author: Steven Xijin Ge
date: '2018-07-02'
slug: forecasting-mass-shootings
categories:
  - R
  - Social issues
tags:
  - Gun violence
header:
  caption: ''
  image: ''
---
   

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Using the mass shooting data from [Mother Jones website,](https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/)
I tried to forecast the number of mass shoottings and total fatalities in the rest of 2018. Some of the R scripts are from this [website.](http://a-little-book-of-r-for-time-series.readthedocs.io/en/latest/src/timeseries.html) Please let me know if I there is any error. I am not an expert on time series analysis. [Twitter](www.twitter.com) @StevenXGe

In recent years, mass shootings are becomming more frequent and more deadly. The models suggest for the rest of 2018, there probably be about 6 more shootings, killing about 59 people in the U.S. 

#Data 
```{r cars}
url = "https://docs.google.com/spreadsheet/pub?key=0AswaDV9q95oZdG5fVGJTS25GQXhSTDFpZXE0RHhUdkE&output=csv"
x = read.csv(url)
colnames(x)
```


#Data cleaning
```{r pressure}
x = x[,c("date","year","Fatalities","Race","Gender","Type")]
x$Race = gsub("white","White",as.character(x$Race) )
x$Race = as.factor(x$Race)
summary(x)
x = subset(x,x$year != 2018)
x$date = as.Date(x$date,"%m/%d/%Y")
```

# The number of mass shootings per year
There is no mass shooting in 1983, 1985, and 2002. 
```{r}
yearly = table(x$year)
yearly = c(1,0, yearly)
names(yearly)[1:4]= 1982:1985
yearly[3]=2
yearly[4]=0 
yearly = c(yearly[1:20],0, yearly[21:35])
names(yearly)[21] = "2002"
barplot(yearly)
```

Clearly, the frequency of mass shootings is increasing in the last 6 years. 



## Time series model
```{r}
series = ts(yearly,start=c(1982)) 
yforecast = HoltWinters(series, beta=FALSE, gamma=FALSE)
yforecast
```

```{r}
plot(yforecast, main="Number of mass shootings ")
```

The red line is probably the model. 

## Making forecast
```{r}
# install.packages("forecast")
library(forecast)
shottings = forecast(yforecast, h = 1)
shottings
```

The model predicts there would be about 8.4 mass shootings this year. Since 2 already happened. There probably would be 6 incidences of mass shooting for the rest of 2018. The 95% confidence interval is between 5 and  11. So it is very likely at least 3 more shootings will occur in 2018.

# Total fatalities in 2018
## Prepare data

```{r}
tem = aggregate(x$Fatalities, by=list(x$year), FUN=sum)
tem2 = rbind( tem[1,], c(1983,0), tem[2,],c(1985,0),tem[3:18,], c(2002, 0), tem[19:33, ] ) 
fatalities = tem2[,2]
names(fatalities) = tem2[,1]
barplot(fatalities, las=2)
```

This data is noisy, but we still see an uptake in total fatalities in the last 10 years. 


```{r}
series = ts(fatalities,start=c(1982))
yforecast = HoltWinters(series, beta=FALSE, gamma=FALSE)
yforecast
```

```{r}
plot(yforecast, main="Total fatalities from mass shooting")
```
```{r}
shottings = forecast(yforecast, h = 1)
shottings
```

So this model predicts that about 80 people will be killed in mass shootings in 2018. Since there were 21 in Florida and Pennsylvania already. The rest of 2018 will see about 59 people getting killed. I hope these are not going to happen in schools. 
